---
title: "Blog post 2"
author: "Kasliwal, Dakshesh; Laz, Thomas; Li, Siyan; Oh, Yunha; Xia, Chuancheng"
date: "2025-03-05"
date-modified: "2025-03-05"
draft: FALSE
---

We chose to analyze the dataset about HIV cases in New York City.

### Data Background

**Where does the data come from? Who collected it? Why was it collected? Are you able to find the data from the original source?**

The data originates from the New York City Department of Health and Mental Hygiene (NYC DOHMH), which collected and compiled information on new HIV and AIDS diagnoses reported through March 31, 2021. Specifically, this dataset covers cases diagnosed from calendar years 2016 through 2020. The data were collected to monitor trends in HIV and AIDS diagnoses, enabling public health officials to track disease patterns, guide prevention and treatment strategies, and identify disparities across different neighborhoods and demographic groups within NYC. Cases are stratified by United Hospital Fund (UHF) neighborhoods, sex, and race/ethnicity, allowing for targeted public health responses. The original dataset can typically be obtained directly from the NYC DOHMH or accessed through official public health data portals maintained by New York City.

**Are there any issues you can see with how the data was collected? What is the sample population? Are there reasons to think the sample is biased in some way?**

The sample population originally includes 8,976 observations. After removing observations that had null entries, the sample size decreased to 7,062. The dataset should be highly accurate, given that the data was collected by a government organization. Additionally, providers and labs are legally required to report any new diagnoses of HIV and AIDS. One way the sample could be biased is that there could be people who don’t get their HIV/AIDS diagnosed, and therefore they wouldn’t be represented in this dataset. This could be a problem if there are a large number of people not reporting their symptoms, and if the people in this group share similar characteristics and/or demographics.

### Data for Equity

Given this is a highly sensitive dataset, it is important to be mindful and intentional when collecting and analyzing data. When looking at the instrumentation of the data, it is important to keep beneficence in mind and “Minimize the amount of personally identifiable information collected.” In this instance, NYC DOHMH asked for the respondent's neighborhood, which is highly granular and potentially sensitive data. To counteract any risks of re-identification, DOHMH doesn’t ask for many other significant demographics, only “borough”, “race/sex”, and “ethnicity”. Additionally, DOHMH lists that one of the reasons for “N/A” cells in the dataset is cell suppression. According to NCES.ed.gov, one of the benefits of cell suppression is “protecting frequency data in tabular form”. However, as mentioned before, the N/A observations do lower our sample size, which could disproportionately affect a certain population or demographic when doing our analysis. 

When looking at the collection of data, “respect for persons” is very important. According to the data description, it seems that the HIV/AIDS data is collected through labs and doctors, not patients themselves. Given this, it is important for there to be some level of consent and awareness of this data collection practice by patients. As we mentioned before, re-identification is a risk when collecting such granular and sensitive data, so it is important to make patients aware and get their consent about these data-collecting practices.

### Data Loading and Cleaning

The .rds file of our clean dataset was created successfully. The data was in a relatively clean state, with three revisions made so far. First, while the data from 2016-2021 has neighborhoods listed as well as the borough, the data from 2010-2013 only had the neighborhood, so the boroughs had to be filled in. We couldn't use the fill function because the data wasn't stored sequentially, so matching each neighborhood to a borough and filling in the boroughs in the csv file itself was the easiest way to do this. Second, all rows with any NA values were removed, using the complete.cases function from R's documentation. Third, the dataset itself stored the Asian/Pacific Islander race in different ways, despite the spelling being the same in the csv file, using a line break instead of a space sometimes, which was fixed using a combination of str_replace_all and str_trim. 

Below is some basic code for a boxplot of the total number of HIV diagnoses, grouped by race.
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(here))

df <- read_rds(here::here("dataset", "hiv_clean.rds"))

ggplot(df, aes(x = `RACE/ETHNICITY`, y = `TOTAL NUMBER OF HIV DIAGNOSES`, fill = `RACE/ETHNICITY`)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  labs(title = "Distribution of HIV Diagnoses by Race/Ethnicity",
       x = "Race/Ethnicity",
       y = "Total Number of HIV Diagnoses") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylim(0, 100)
```
Below is code for a bar chart, grouped by race.
```{r}
ggplot(df, aes(x = reorder(`RACE/ETHNICITY`, -`TOTAL NUMBER OF HIV DIAGNOSES`), 
                        y = `TOTAL NUMBER OF HIV DIAGNOSES`, fill = `RACE/ETHNICITY`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Total HIV Diagnoses by Race/Ethnicity",
       x = "Race/Ethnicity",
       y = "Total Number of HIV Diagnoses") +
  theme_minimal() +
  theme(legend.position = "none")
```

Based on these graphs, some trends noticed are higher levels of cases for Black and Hispanic people, compared to other races, despite lower populations, indicating potential underlying causes of this discrepancy. The "races" all, other/unknown, and unknown will be removed in further analysis.