---
title: "Blog Post 4"
author: ""
date: "2025-03-28"
date-modified: "2025-03-28"
draft: FALSE
---

From “Line Charts: Trends Over Time (data.qmd)”, the increase is observed only in the Asian/Pacific Islander category, even though a general reduction in HIV diagnoses rates over time across all racial/ethnic groups. 

##Comparing "Total Number of HIV Diagnoses" by Sex for Asian/Pacific Islanders
```{r}
suppressPackageStartupMessages(library(tidyverse))
library(readr)
library(here)

df <- read_rds(here::here("dataset", "hiv_clean.rds"))


ggplot(df, aes(x = SEX, y = as.numeric(`TOTAL NUMBER OF HIV DIAGNOSES`), fill = SEX)) +
  geom_bar(stat = "identity") +
  labs(title = "Total HIV Diagnoses for Asian/Pacific Islanders by Sex",
       x = "Sex",
       y = "Total Number of HIV Diagnoses") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar graph clearly indicates a significant gender disparity in HIV diagnoses among Asian/Pacific Islanders, with males accounting for a substantially higher number of cases compared to females. The number of diagnoses among males is several times higher than that among females, highlighting a pronounced vulnerability or increased risk exposure for males in this population.


##Line Graph of HIV Diagnoses Rates per 100,000 Population Over Years by Sex
```{r}
ggplot(df, aes(x = YEAR, y = as.numeric(`HIV DIAGNOSES PER 100,000 POPULATION`), color = SEX)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "HIV Diagnoses Rates (per 100,000) for Asian/Pacific Islanders by Sex",
       x = "Year",
       y = "HIV Diagnoses per 100,000 Population") +
  theme_minimal() +
  scale_x_continuous(breaks = unique(df$YEAR))

```

The rate of HIV diagnoses for males (blue dots) is consistently higher than for females (green dots) across all years. Noticeably high spikes occur, especially around 2010-2013 and again in 2017-2018.

